### 2023-11-20 submission/response_to_reviewers.Rmd

**1. I found the result of increasing CoV in Bray-Curtis distances with increasing sequencing depth to be an interesting result given that increased sequencing depth should generally result in better confidence of what is within the underlying population. I'm wondering if the CoV results are difficult to understand due to the changing number of samples included at each sampling depth. Could this analysis be redone using only samples with sequencing depths that are above the highest rarefaction value chosen so that sample number does not vary across depth (this could be done in a subset of the samples/datasets presented). If changing sample number is not to blame are there any further speculations as to what is going on here?**

> I have added a set of panels to Figure 6 indicating the standard deviation across the samples. This hopefully makes it clear that the standard deviation between samples is decreasing, but not as fast as the mean distance between samples. On the whole, the COV values are fairly small. I have added text to this effect at LXXX.

**2. It is clear based on the results shown that treatment groups that are highly associated with variance in sequencing depth will result in high false positive rates during diversity analysis when rarefaction is not used. However, I would argue that the simulation used does not in general reflect many natural datasets. It would be uncommon for most natural datasets to vary by treatment group by splitting sequencing depth via the median. I am wondering if this analysis could be done with varying levels of association between treatment group and sequencing depth. This will give the readers a better idea of how large this issue is in actual practice as the true association for most datasets may lay somewhere between figures 2 and 3. This would also help address how much of a concern this may be for previously published papers that do not use rarefaction during diversity analysis.**

> I appreciated this suggestion and repeated the analysis with a less severe level of confounding. In addition to the perfect confounding of treatment groups by sequencing effort, I created an additional simulation where the smallest 5% of samples were assigned to one treatment group and the largest were assigned to another; the remaining samples were randomly assigned between the groups. The new results are in Figure 3 and the previous simulation are now in Figure S4. Although the results are less severe in the new simulation, they still affirm the effects of sampling effort on the propensity to get false positive results.

### 2023-11-20 submission/manuscript.Rmd
Implement less extreme confounding analysis

***Rarefaction limits the detection of false positives when sequencing effort and treatment group are confounded.*** Next, I investigated the impact of the various strategies and metrics on falsely detecting a significant difference using the the same communities generated from the null model in the analysis of alpha and beta diversity metrics. To test for differences in alpha and beta diversity I used the non-parametric Wilcoxon test and non-parametric permutation-based multivariate analysis of variance (PERMANOVA). First, I employed an unbiased null treatment model to measure the false detection rate, which should not have meaningfully differed from 5%. Indeed, for each dataset and alpha and beta diversity metric and strategy for accounting for uneven sequencing, approximately 5% of the tests yielded a significant result (Figure 2). Second, I employed a biased null treatment model where the treatment group was completely confounded with the number of sequences in each sample. Under this model, only the data normalized with rarefaction consistently resulted in a 5% false positive rate for alpha and beta diversity metrics (Figure 2). These results aligned with the observed sensitivity of alpha and beta diversity metrics to sequencing effort and underscore the value of rarefaction.

@@

**Null treatment models.** I created an unbiased and biased treatment model. In the unbiased model, samples were randomly assigned to one of two treatment groups. In the biased treatment model, samples that had more than the median number of sequences for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group. Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\alpha$) of 0.05.

@@

**Figure 3.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth is confounded with treatment group**. Samples were assigned to different treatment groups based on whether they were above the median number of sequences for each dataset. To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold. The false positive rate was the number of times a dataset yielded a significant result.

@@

>>>>

***Rarefaction limits the detection of false positives when sequencing effort and treatment group are confounded.*** Next, I investigated the impact of the various strategies and metrics on falsely detecting a significant difference using the the same communities generated from the null model in the analysis of alpha and beta diversity metrics. To test for differences in alpha and beta diversity I used the non-parametric Wilcoxon test and non-parametric permutation-based multivariate analysis of variance (PERMANOVA). First, I employed an unbiased null treatment model to measure the false detection rate, which should not have meaningfully differed from 5%. Indeed, for each dataset and alpha and beta diversity metric and strategy for accounting for uneven sequencing, approximately 5% of the tests yielded a significant result (Figure 2). Second, I employed a biased null treatment model where the treatment group was completely confounded with the number of sequences in each sample (Figure S4). Under this model, only the data normalized with rarefaction consistently resulted in a 5% false positive rate for alpha and beta diversity metrics. I then used a less severe bias where the smallest 5% of samples were assigned to one treatment group and the largest 5% were assigned to another; the other samples were randomly assigned to both treatment groups. Although the sensitivity of the false positive rate to the level of confounding varied across methods of quantifying alpha and beta diversity, rarefaction was the only method to consistently control the false positive rate (Figure 3). These results aligned with the observed sensitivity of alpha and beta diversity metrics to sequencing effort and underscore the value of rarefaction.

@@

**Null treatment models.** I created an unbiased and biased treatment model. In the unbiased model, samples were randomly assigned to one of two treatment groups. I created two biased treatment models. In the first, samples that had more than the median number of sequences for a dataset were assigned to one treatment group and the rest were assigned to a second treatment group (see Figure S4). In the second, samples whose sequencing depth where in the 5th percentile or lower were assigned to one treatment group while those in the 95% percentile or higher were assigned to another; the remaining samples were randomly assigned to either treatment group (see Figure 3). Comparison of any diversity metric across the two treatment groups should have only yielded a significant result in 5% of the simulations when testing under a Type I error (i.e., $\alpha$) of 0.05.

@@

**Figure 3.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth is confounded with treatment group**. Samples were assigned to different treatment groups so that the smallest 5% of samples were assigned to one treatment group and the largest 5% to another; the other 90% of samples were randomly assigned to both treatment groups. To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold. The false positive rate was the number of times a dataset yielded a significant result.

@@

\includegraphics[height=17cm]{figure_s4.png}

**Figure S4.The risk of falsely detecting a difference between treatment groups drawn from a null model does not meaningfully vary from 5% when data are normalized by rarefaction when sequencing depth is confounded with treatment group**. Samples were assigned to different treatment groups based on whether they were above the median number of sequences for each dataset. To calculate the false detection rate, datasets were regenerated 100 times and differences in alpha diversity were tested using a Wilcoxon test (A) and differences in beta diversity were tested using PERMANOVA (B) at a 5% threshold. The false positive rate was the number of times a dataset yielded a significant result.

### 2023-11-20 submission/manuscript.tex
Implement less extreme confounding analysis

\textbf{\emph{Rarefaction limits the detection of false positives when
sequencing effort and treatment group are confounded.}} Next, I
investigated the impact of the various strategies and metrics on falsely
detecting a significant difference using the the same communities
generated from the null model in the analysis of alpha and beta
diversity metrics. To test for differences in alpha and beta diversity I
used the non-parametric Wilcoxon test and non-parametric
permutation-based multivariate analysis of variance (PERMANOVA). First,
I employed an unbiased null treatment model to measure the false
detection rate, which should not have meaningfully differed from 5\%.
Indeed, for each dataset and alpha and beta diversity metric and
strategy for accounting for uneven sequencing, approximately 5\% of the
tests yielded a significant result (Figure 2). Second, I employed a
biased null treatment model where the treatment group was completely
confounded with the number of sequences in each sample. Under this
model, only the data normalized with rarefaction consistently resulted
in a 5\% false positive rate for alpha and beta diversity metrics
(Figure 2). These results aligned with the observed sensitivity of alpha
and beta diversity metrics to sequencing effort and underscore the value
of rarefaction.

@@

\textbf{\emph{Increased rarefaction depth reduces intra-sample variation
in alpha and beta diversity.}} Once concern with using rarefaction is
the perceived loss of sequencing information when a large fraction of
data appears to be removed when the community with the greatest
sequencing depth is sampled to the size of the community with the least
(e.g., the smallest sample in the Bioethanol dataset had 1.04\% of
sequences that were in the largest sample). To assess the sensitivity of
alpha and beta diversity metrics to rarefaction depth, I again used the
dataset generated using the null models, but used rarefaction with each
community to varying sequencing depths (Figure 6). The richness values
increased with sequencing effort as rare OTUs would continue be
detected. In contrast, the Shannon diversity and Bray-Curtis values
plateaued with increased sequencing effort. This result was expected
since increased sequencing would lead to increased precision in the
measured relative abundance of the OTUs. Next, I measured the
coefficient of variation (i.e., the mean divided by the standard
deviation) between samples for richness, Shannon diversity, and
Bray-Curtis distances. Although the mean richness appeared to increased
unbounded with sequencing effort, the coefficient of variation for each
dataset was relatively stable. In general, the coefficient of variation
increased slightly with sequencing depth only to decline once smaller
samples were removed from the analysis at higher sequencing depths.
Interestingly, the coefficient of variation between Shannon diversity
values decreased towards zero with increased sequencing effort and the
coefficient of variation between Bray-Curtis distances tended to
increase. Regardless, the coefficients of variation were relatively
small. This analysis indicates that there are benefits to increased
sequencing depths.

@@

The up to 100-fold difference in sample sizes is an unfortunate
byproduct of how sequencing libraries are constructed. Researchers
perform separate PCRs for each sample with unique index (aka, barcode)
sequences that allow them to later assign sequences back to the samples
that they came from. When the PCR products are pooled, efforts are taken
to pool the fragments in equimolar ratios. Researchers use one of two
approaches. First, they often will quantify the concentration of DNA
from each PCR and then pool DNA in the desired amounts. Alternatively,
they may use normalization plates where each well can hybridize a
uniform amount of DNA that is then eluted and pooled. Clearly, both
approaches have limitations that reduce the ability to truly achieve
equimolar mixture. In addition, for some samples it is common to
co-amplify non-specific DNAs which may add to the challenges of
achieving equimolar mixtures of the desired amplicons since those
sequences will be removed (28). Regardless, it is clear that better
strategies are needed to reduce the variation in the number of sequences
generated for each sample.

@@

\textbf{Null treatment models.} I created an unbiased and biased
treatment model. In the unbiased model, samples were randomly assigned
to one of two treatment groups. In the biased treatment model, samples
that had more than the median number of sequences for a dataset were
assigned to one treatment group and the rest were assigned to a second
treatment group. Comparison of any diversity metric across the two
treatment groups should have only yielded a significant result in 5\% of
the simulations when testing under a Type I error (i.e., \(\alpha\)) of
0.05.

@@

\textbf{Figure 3.The risk of falsely detecting a difference between
treatment groups drawn from a null model does not meaningfully vary from
5\% when data are normalized by rarefaction when sequencing depth is
confounded with treatment group}. Samples were assigned to different
treatment groups based on whether they were above the median number of
sequences for each dataset. To calculate the false detection rate,
datasets were regenerated 100 times and differences in alpha diversity
were tested using a Wilcoxon test (A) and differences in beta diversity
were tested using PERMANOVA (B) at a 5\% threshold. The false positive
rate was the number of times a dataset yielded a significant result.

@@

>>>>

\textbf{\emph{Rarefaction limits the detection of false positives when
sequencing effort and treatment group are confounded.}} Next, I
investigated the impact of the various strategies and metrics on falsely
detecting a significant difference using the the same communities
generated from the null model in the analysis of alpha and beta
diversity metrics. To test for differences in alpha and beta diversity I
used the non-parametric Wilcoxon test and non-parametric
permutation-based multivariate analysis of variance (PERMANOVA). First,
I employed an unbiased null treatment model to measure the false
detection rate, which should not have meaningfully differed from 5\%.
Indeed, for each dataset and alpha and beta diversity metric and
strategy for accounting for uneven sequencing, approximately 5\% of the
tests yielded a significant result (Figure 2). Second, I employed a
biased null treatment model where the treatment group was completely
confounded with the number of sequences in each sample (Figure S4).
Under this model, only the data normalized with rarefaction consistently
resulted in a 5\% false positive rate for alpha and beta diversity
metrics. I then used a less severe bias where the smallest 5\% of
samples were assigned to one treatment group and the largest 5\% were
assigned to another; the other samples were randomly assigned to both
treatment groups. Although the sensitivity of the false positive rate to
the level of confounding varied across methods of quantifying alpha and
beta diversity, rarefaction was the only method to consistently control
the false positive rate (Figure 3). These results aligned with the
observed sensitivity of alpha and beta diversity metrics to sequencing
effort and underscore the value of rarefaction.

@@

\textbf{\emph{Increased rarefaction depth reduces intra-sample variation
in alpha and beta diversity.}} Once concern with using rarefaction is
the perceived loss of sequencing information when a large fraction of
data appears to be removed when the community with the greatest
sequencing depth is sampled to the size of the community with the least
(e.g., the smallest sample in the Bioethanol dataset had 1.04\% of
sequences that were in the largest sample). To assess the sensitivity of
alpha and beta diversity metrics to rarefaction depth, I again used the
dataset generated using the null models, but used rarefaction with each
community to varying sequencing depths (Figure 6). The richness values
increased with sequencing effort as rare OTUs would continue be
detected. In contrast, the Shannon diversity and Bray-Curtis values
plateaued with increased sequencing effort. This result was expected
since increased sequencing would lead to increased precision in the
measured relative abundance of the OTUs. Next, I measured the
coefficient of variation (i.e., the mean divided by the standard
deviation) between samples for richness, Shannon diversity, and
Bray-Curtis distances. Although the mean richness appeared to increased
unbounded with sequencing effort, the coefficient of variation for each
dataset was relatively stable. In general, the coefficient of variation
increased slightly with sequencing depth only to decline once smaller
samples were removed from the analysis at higher sequencing depths.
Interestingly, the coefficient of variation between Shannon diversity
values decreased towards zero with increased sequencing effort and the
coefficient of variation between Bray-Curtis distances tended to
increase slightly for samples where the standard deviation did not
decrease as quickly as the mean distance between samples. Regardless,
the coefficients of variation were relatively small. This analysis
indicates that there are benefits to increased sequencing depths.

@@

The up to 100-fold difference in sample sizes is an unfortunate
byproduct of how sequencing libraries are constructed. Researchers
perform separate PCRs for each sample with unique index (aka, barcode)
sequences that allow them to later assign sequences back to the samples
that they came from. When the PCR products are pooled, efforts are taken
to pool the fragments in equimolar ratios. Researchers use one of two
approaches. First, they often will quantify the concentration of DNA
from each PCR and then pool DNA in the desired amounts. Alternatively,
they may use normalization plates where each well can hybridize a
uniform amount of DNA that is then eluted and pooled. Clearly, both
approaches have limitations that reduce the ability to truly achieve
equimolar mixture. In addition, for some samples it is common to
co-amplify non-specific DNAs which may add to the challenges of
achieving equimolar mixtures of the desired amplicons since those
sequences will be removed (28). Regardless, it is clear that better
strategies are needed to reduce the variation in the number of sequences
generated for each sample. Furthermore, it is important to keep in mind
that because of residual sequencing errors, PCR artifacts, and
incomplete sequencing, it is not possible to obtain absolute
quantification of alpha or beta diversity values. Instead, comparisons
of these values need to be made on a relative basis.

@@

\textbf{Null treatment models.} I created an unbiased and biased
treatment model. In the unbiased model, samples were randomly assigned
to one of two treatment groups. I created two biased treatment models.
In the first, samples that had more than the median number of sequences
for a dataset were assigned to one treatment group and the rest were
assigned to a second treatment group (see Figure S4). In the second,
samples whose sequencing depth where in the 5th percentile or lower were
assigned to one treatment group while those in the 95\% percentile or
higher were assigned to another; the remaining samples were randomly
assigned to either treatment group (see Figure 3). Comparison of any
diversity metric across the two treatment groups should have only
yielded a significant result in 5\% of the simulations when testing
under a Type I error (i.e., \(\alpha\)) of 0.05.

@@

\textbf{Figure 3.The risk of falsely detecting a difference between
treatment groups drawn from a null model does not meaningfully vary from
5\% when data are normalized by rarefaction when sequencing depth is
confounded with treatment group}. Samples were assigned to different
treatment groups so that the smallest 5\% of samples were assigned to
one treatment group and the largest 5\% to another; the other 90\% of
samples were randomly assigned to both treatment groups. To calculate
the false detection rate, datasets were regenerated 100 times and
differences in alpha diversity were tested using a Wilcoxon test (A) and
differences in beta diversity were tested using PERMANOVA (B) at a 5\%
threshold. The false positive rate was the number of times a dataset
yielded a significant result.

@@

\includegraphics[height=17cm]{figure_s4.png}

\textbf{Figure S4.The risk of falsely detecting a difference between
treatment groups drawn from a null model does not meaningfully vary from
5\% when data are normalized by rarefaction when sequencing depth is
confounded with treatment group}. Samples were assigned to different
treatment groups based on whether they were above the median number of
sequences for each dataset. To calculate the false detection rate,
datasets were regenerated 100 times and differences in alpha diversity
were tested using a Wilcoxon test (A) and differences in beta diversity
were tested using PERMANOVA (B) at a 5\% threshold. The false positive
rate was the number of times a dataset yielded a significant result.
